Tackling discrepancies in freshwater invertebrate trait databases and taxonomic resolution: Harmonising across continents and trait aggregation

### 2021-01-22 Paper/Draft_harmonization_paper.tex
22.01: Added (final) comments and corrections from Co-Authors

1. Invertebrate traits are expected to be suitable for large-scale analysis in freshwater ecology due to their consistent response to environmental gradients. In recent years, comprehensive invertebrate trait databases have been established at different scales (e.g. regions, continents). The wide availability of invertebrate trait data supports trait-based studies, especially on large scales. However, a number of data-related problems hamper the work with trait data. For example, for freshwater invertebrate traits standardised definitions are lacking impeding trait data synthesis. Moreover, uncertainty remains how harmonising grouping features from different trait databases can influence trait-environment relationships. Another issue is that if the taxonomic resolution of observational taxonomic data is less precise than in trait databases, the traits in databases require aggregation before they can be linked to the observational data.
% Here: Describe aims of the study or redundant with part 2? Something like this: The aim of our study is to investigate these data-related issues by synthesizing trait data from different databases.

2. We describe discrepancies in trait definitions of trait databases from Europe, North America, New Zealand, and Australia for traits frequently used in ecological studies. Based on these trait databases, we established 4 novel trait datasets by harmonising 7 commonly used grouping features. Two of these datasets were used to evaluate different trait aggregation methods by comparing aggregated traits with assigned traits by experts at family-level. We tested trait aggregation based on the mean and based on the median, each with different weightings. Furthermore, we explored the effect of grouping feature harmonisation and trait aggregation by re-analysing data from a case study.

3. Between the trait databases, grouping features differed in terms of their classification into traits and their scope leading to discrepancies how traits are defined. Also, the coding to describe traits varied between databases. Aggregated and assigned traits based on expert knowledge displayed a good match, specially approaches using the median. Grouping feature harmonisation and trait aggregation only slightly changed the results in the re-analysis compared to the original study, although averaging effects can occur due to harmonisation.

@@

Explaining and predicting how communities are shaped by environmental factors is one of the main goals of ecology. Species traits, defined as measurable properties of an organism (\cite{mcgill_rebuilding_2006}), might be beneficial in achieving this goal (\cite{heino_jani_macroecological_2013}). Traits evolve through adaptations (e.g., physiological, behavioural, etc.) of organisms to their environment and indicate direct or indirect linkages between the biological response of an organism or a population and its environment (\cite{southwood_habitat_1977, verberk_delivering_2013}).
Besides providing a mechanistic explanation of species-environment relationships, trait-based approaches might be suitable for large-scale analysis because the variability in trait responses is lower than in taxonomic responses (\cite{bonada_taxonomic_2007, baird_toward_2011}). Traits of individual freshwater invertebrate are difficult to determine because - unlike plant traits - they are often difficult to measure directly. For example, to gain knowledge on feeding habits requires evaluating mouthpart morphology, consumed food, and the organisms function within its community (\cite{moog_comprehensive_nodate}). Nevertheless, invertebrate traits have been increasingly used in freshwater ecology, e.g. by relating macroinvertebrate trait composition to environmental factors and as trait metrics in biomonitoring (\cite{bhowmik_large_2015, menezes_beyond_2010, poff_developing_2010, szocs_effects_2014}).

In the last decades, freshwater ecologists have compiled comprehensive invertebrate trait databases for various continents (\cite{usseglio-polatera_biomonitoring_2000, schmidt-kloiber_www.freshwaterecology.info_2015, vieira_database_nodate, Philips_and_Smith_NZ_DB_2018, kefford_integrated_2020, tomanova_trophic_2006}). The availability of invertebrate trait data from different continents enables comparisons of trait variation and analyses of their relation to environmental factors across large scales. To date, such analyses have been carried out mostly within continents, using information from one or two trait databases. For example, \citet{bonada_taxonomic_2007} compared trait composition for Mediterranean and temperate regions in Europe using traits from \citet{ussegliopolatera_biological_2000} (typically referred to as Tachet database); \citet{poff_developing_2010} characterized trait composition across sites in the Western US using traits from \citet{poff_functional_2006}, and \citet{botwe_effects_2018} investigated the effect of salinity on invertebrate traits in different sites in South Australia using trait data from \citet{poff_functional_2006} and \citet{schafer_trait_2011}. Analyses of invertebrate traits that synthesize information on invertebrate grouping features from more than two different continents are rare.

@@

Traits can be described in a binary fashion, or with multiple categories, ignoring uncertainty how the trait is expressed in any particular organism (e.g. adult terrestrial stage, presence of gills) or continuous (e.g. tolerance of pollution, body size). One approach for dealing with uncertainty is the use of fuzzy coding, where the traits are assigned probabilistic values. Fuzzy codes are used to account for plasticity in traits, variability in traits within taxonomic groups above species, and incomplete knowledge and are usually converted to proportions. In this study we denote the probabilistic values (0-1) of traits as affinities or affinity scores.

However, invertebrate trait databases are heterogeneous regarding the coding they use for their traits (\cite{culp_incorporating_2011}). For example, in the study from \citet{brown_functional_2018} the authors needed to reclassify traits using expert knowledge given that the individual databases employed different coding approaches (i.e. European and New Zealand databases fuzzy coding, North American database binary coding). Databases with standardised and unambiguous traits and a consistent coding of traits would minimize the data processing effort. Discrepancies in the taxonomic resolutions (e.g. species, genus or family) when linking observational taxonomic data to trait database represents another challenge. When observations are on a more precise taxonomic level than data available in the trait databases (e.g. observations on species-level, trait data on genus-level) trait information of the less precise taxonomic level is often assigned (e.g. \cite{szocs_effects_2014, vos_taxonomic_2017}). Conversely, if trait information is only available on more precise taxonomic levels than the observed taxa, traits are aggregated to a less precise taxonomic level (e.g. \cite{aspin_extreme_2019, piliere_a._f._h._importance_2016, poff_functional_2006, szocs_effects_2014}). Studies have used different methods of trait aggregation, e.g. the mean (\cite{magliozzi_functional_2019}), median (\cite{szocs_effects_2014}) or the mode (\cite{piliere_a._f._h._importance_2016}). Studies on how and to which extent different trait aggregation methods influence trait-based analysis are missing. However, related knowledge would inform future studies regarding the choice of the aggregation method.

The outlined difficulties when working with or synthesizing trait data motivated us to (1) investigate discrepancies in trait definition between trait databases from Europe, North America, Australia, and New Zealand. As a basis for further evaluating different trait aggregation methods and how harmonising definition discrepancies between grouping features can influence trait-environment relationships, we established 4 novel invertebrate trait datasets from these trait databases by harmonising 7 grouping features. We evaluate different trait aggregation methods by (2) comparing aggregated traits with assigned traits by experts at family-level for two of the established trait datasets. (3) we re-analysed data on the effect of anthropogenic salinisation on biological traits by \citet{szocs_effects_2014} using trait data from the established harmonised European trait datasets and aggregated traits. By comparison with the original analysis, we investigated how harmonising and aggregating trait data can modify the outcome of trait-environment relationships.

@@

We extracted information from 6 trait databases from Europe, North America, Australia, and New Zealand and harmonised 7 grouping features. Trait information for Europe was obtained from the freshwaterecology.info database (\cite{schmidt-kloiber_www.freshwaterecology.info_2015}) and the Tachet database (\cite{ussegliopolatera_biological_2000}). The freshwaterecology.info contains taxa on species-level, while taxa recorded in the Tachet database are on species, genus and family-level. Trait information for North America was obtained from \citet{twardochleb_trait_data_2020} and complemented by \citet{vieira_database_nodate}. Data on body form for European and North American taxa was based on expert knowledge (\cite{polatera_personal_information_2020}). For Australia and New Zealand, we used trait databases from Kefford et al. (\cite{kefford_integrated_2020}) and Philips and Smith respectively (\cite{Philips_and_Smith_NZ_DB_2018}). To increase readability we refer to the databases as well as the datasets we extracted from them by the name of the continent they originate from, except for European databases, which we refer to by their commonly used names (freshwaterecology.info and Tachet database).

We selected traits of seven grouping features that were available in all databases, are commonly applied in trait-based ecological studies, and describe different parts of the biology of a species: life history (Voltinism), morphology (Respiration, Body form, Size), ecology (Locomotion, Feeding mode) and reproduction (Oviposition). We omitted ecological traits that describe habitat preferences (e.g. temperature preference) because these traits are missing in the New Zealand trait database. The grouping features were differently classified across the databases, we therefore harmonised them into 26 traits (Table \ref{tab:traits_harmonisation}). Harmonisation was undertaken by amalgamating similar traits into one trait (e.g. crawlers and sprawlers into crawlers). Thereby, for a particular taxon the highest trait affinity score among the amalgamated traits was taken.

We used fuzzy coding for establishing our harmonised datasets unless data quality prohibited. In the latter case we used binary traits, i.e. categorical and continuous traits were converted into binary traits. For example, in the  freshwaterecology.info database, the classification of the trait voltinism accounts for different faunistic regions. Hence, entries such as "arctic" or "boreal" in the voltinism traits were substituted with a value of 1. Implicitly, we assumed for binary traits that a value of 1 and 0 corresponds to the highest and no affinity for a particular trait. Fuzzy codes are reported with different ranges in the trait databases (e.g. freshwaterecology.info 0 to 10, Tachet database 0 to 3 or 0 to 5). We standardised these to a range between 0 and 1 and converted trait affinities to percentages. Thus, fuzzy coded and binary traits were in the same range.

Prior to harmonisation we consolidated duplicate taxa on species, genus or family-level if present within a dataset, by either applying the median for fuzzy coded traits, or the maximum for binary traits. We omitted taxa with a lower taxonomic precision than family-level.

@@

    \textit{a} The trait "bluff (blocky)" occurred in the North America (Vieira et al.) database and was newly classified by expert knowledge into cylindrical and flattened (\cite{polatera_personal_information_2020}). \\  

@@

Aggregated trait affinities of the five trait aggregation methods (\textit{direct\_agg\textsubscript{median}}, \textit{direct\_agg\textsubscript{mean}}, \textit{stepwise\_agg\textsubscript{median}}, \textit{stepwise\_agg\textsubscript{mean}}, and \textit{weighted\_agg}) were compared to trait affinities assigned at family-level by experts, which were available for Australia and North America for a subset of grouping features and taxa. For the Australian dataset, we compared aggregated trait affinities with assigned trait affinities resolved at family-level for the grouping features feeding mode and size using data from \citet{chessman_dissolved-oxygen_2018}. In \citet{chessman_dissolved-oxygen_2018} feeding mode is classified similarly as in the Australian dataset except that the trait parasite is missing. We conducted the comparison for the 220 families available in \citet{chessman_dissolved-oxygen_2018}. Considering each factor combination of family and trait (in total 8) this amounts to 1760 cases.

@@

\subsection*{Analysis of the effect of harmonisation and trait aggregation on trait-environment relationships}

We repeated the analysis of \citet{szocs_effects_2014} who studied the effect of anthropogenic salinization on invertebrates in the River Werra in Germany. For the re-analysis we used the established harmonised grouping feature dataset for Europe and additionally aggregated traits using the aforementioned aggregation methods.

The river Werra has been subject to effluents from the potash industry since the mid of the 20th century and allows to study responses of invertebrates and their trait compositions to salinization (\cite{bathe_biological_2011}). Sites downstream, upstream, and close to the salt discharge (transition) were compared regarding their trait composition. Further details can be found in \citet{szocs_effects_2014}.

We substituted 6 of the grouping features from the original data with harmonised grouping features from the European trait dataset. We compared the results of the redundancy analysis (RDA) from the original study to the case when using harmonised grouping features. Specifically, the trait composition expressed as community weighted mean (CWM) traits was ordinated along an electric conductivity gradient. We compared the species scores obtained from the RDA, i.e. the coordinates of the tips of the vectors representing the CWM traits in the bi- or triplots. Following the original study, we identified traits associated with high or low salinity based on their distance to the ordination axis median using the Mahalanobis distance. CWM traits with a Mahalanobis distance greater than the $97.5 \%$ -quantile of the Chi-square distribution (5.02) were regarded as traits responding to either low or high salinity. For our analysis, we used the same 21 grouping features that \citet{szocs_effects_2014} used. The 6 harmonised grouping features used were \textit{Size, Feeding mode, Locomotion, Oviposition, Respiration}, and \textit{Voltinism}. Additionally, for testing the effect of aggregated traits we assigned to each taxon in \citet{szocs_effects_2014} the aggregated trait value for its corresponding family and repeated the RDA.

@@

The data processing and aforementioned analyses were carried out using R (Version 3.6.1). Raw data and the R code for data processing and grouping feature harmonisation is located in the Github repository: \url{https://github.com/KunzstLD/Invertebrate_traits}. Scripts and data to reproduce the trait aggregation and analysis with aggregated traits are located in the Github repository \url{https://github.com/KunzstLD/Trait-aggregation}.

@@

Regarding the taxonomic coverage, the New Zealand dataset has the smallest taxon pool (478 taxa, Table \ref{tab:tax_coverage}). By contrast, the European trait dataset has the largest taxon pool with 4601 taxa followed by the North American trait dataset that contained trait information on 3753 taxa. The Australian dataset contains 1402 taxa. The European, New Zealand, and North American datasets have most taxa on the species-level whereas the Australian dataset has a similar number of taxa on species and genus-level.

\begin{table}[ht]
    \centering
    \caption{Number of taxa per harmonised dataset and per taxonomic level. Numbers in parenthesis show rounded relative frequencies in percentage.} 
    \label{tab:tax_coverage}
    \begin{tabular}{lccccc}
    \toprule[.1em]
    Dataset & Nr. of taxa & Species & Genus & Family & Nr. aquatic insects \\ 
    \toprule[.1em]
    EU & 4601 & 3739 (81) & 704 (15) & 158 (3) & 3942 (86) \\ 
    NOA & 3753 & 2414 (64) & 1163 (31) & 176 (5) & 3305 (88) \\ 
    AUS & 1402 & 564 (40) & 578 (41) & 260 (19) & 1016 (72) \\ 
    NZ & 478 & 404 (85) & 47 (10) & 27 (6) & 443 (93) \\ 
    \bottomrule
    \end{tabular}
\end{table}

@@

The amount of entries with available information for the selected grouping features varied strongly for the established European, North American, and Australian trait datasets (Table \ref{tab:trait_coverage}). By contrast, the New Zealand dataset contained complete trait information for most of the investigated grouping features (between 94 \% and 100 \%).

\begin{table}[ht]
    \centering
    \caption{Rounded percentage of entries that have 
    information for the individual grouping features
    shown per trait dataset.} 
    \label{tab:trait_coverage}
    \begin{tabular}{lccccccc}
    \toprule[.1em]
    Dataset & Body form & Oviposition & Voltinism & Locomotion & Size & Respiration & Feeding mode \\ 
    \toprule[.1em]
    EU & 8 & 15 & 23 & 36 & 11 & 57 & 76 \\ 
    NOA & 28 & 13 & 47 & 52 & 73 & 44 & 63 \\ 
    AUS & 4 & 46 & 49 & 39 & 75 & 68 & 99 \\ 
    NZ & 100 & 94 & 100 & 99 & 100 & 100 & 99 \\ 
    \bottomrule
    \end{tabular}
\end{table}

@@

\subsection*{Discrepancies of invertebrate trait definitions}

Definitions of grouping features and traits varied in their level of detail in the investigated trait databases. The freshwaterecology.info database, the Tachet database and the North American database (Twardochleb et al.) provided more detailed descriptions of their trait information compared to the North American (Vieira et al.) and New Zealand database. An exception is the Australian trait database (\cite{kefford_integrated_2020}) which is a collection of seven specific trait datasets. Thus, grouping features occur multiple times with varying differentiation into traits. Depending on the dataset trait information is described with more or less detail.

The definition of grouping features varied across databases mainly concerning their differentiation into traits but also in their scope. We provide a summary of discrepancies in trait definitions in the appendix (Table \ref{stab:trait_definitions}). Both, differences in differentiation and scope can lead to discrepancies in trait definitions. For example, for the grouping feature feeding mode discrepancies arise because traits are assigned in different ways. The Tachet database defines predators as carvers, engulfers and swallowers. By contrast, in the North American (Twardochleb et al.) database predators are defined as engulfers and carnivorous piercers. In turn, in the Tachet database, piercers are defined as a separate trait encompassing herbivorous and carnivorous piercers. Furthermore, the scope in the freshwaterecology.info database for feeding mode is primarily on the food source of a species (except for filterers), while the other databases focus on the strategies of food acquisition. Therefore, the freshwaterecology.info database defines e.g. predator as "eating from prey", while the other databases use the mouthpart morphology as basis of their definition. The Tachet database captures the food source in an additional grouping feature. Varying levels of differentiation are also present in all other investigated grouping features between the trait databases (Table \ref{tab:trait_databases_coding_differentiation} and Table \ref{stab:trait_definitions}). Locomotion definitions differ also in scope between databases. Freshwaterecology.info and New Zealand databases describe locomotion as the way of movement of an organism, Tachet as substrate relation and locomotion, the North American (Vieira et al.) as how organisms deal with flow, Australia as attachment, and the North American (Twardochleb et al.) database includes not only the way of movement, but also the location of movement. Similarly, regarding the reproduction traits, databases differ in their scope. Reproduction is captured in one grouping feature and defined as location of oviposit clutches and mode of reproduction in the freshwaterecology.info and Tachet databases. North America (Vieira et al.) provides information on the oviposition location but not on reproductive behavior. The Australian database report traits for reproductive behavior but also on oviposition site. The New Zealand database distinguishes three grouping features related to reproduction: reproductive technique, oviposition site (e.g. water surface, terrestrial), and egg/egg mass (e.g. free, cemented).

Various codings to describe traits are used throughout the databases (e.g. binary, fuzzy, continuous). The freshwaterecology.info and Australian use different codings in their databases. Tachet and the New Zealand database exclusively use fuzzy coding. Both North American trait databases contain categorical grouping features that can be converted into traits using a binary coding (Table \ref{tab:trait_databases_coding_differentiation}).

@@

We evaluate the simulation results based on the produced range of aggregated trait affinities and by comparing results of every simulated dataset.

As expected with increasing trait variability the produced range of trait affinities increased for all aggregation methods (Figure \ref{fig:overview_sim_results}).% Ben would add: 
 % as indicated by the increasing range of the boxes and wiskers with increasing standard deviation.

>>>>

1. Use of invertebrate traits rather than species composition is expected  expected to facilitate large-scale comparisons of community structure and responses to disturbance in freshwater ecology because the same traits can potentially occur everywhere. % OLD: because of their consistent response to environmental gradients
In recent years, comprehensive invertebrate trait databases have been established at different scales (e.g. regions, continents). The wide availability of invertebrate trait data supports trait-based studies, especially at large scales. However, a number of data-related problems hamper the work with trait data. For example, for freshwater invertebrate traits standardised definitions are lacking, impeding comparisons across regions. Moreover, it is uncertain how standardising grouping features from different databases might influence trait-environment relationships. Another issue is that if the taxonomic resolution of observational taxonomic data is less precise than in trait databases, the traits in databases require aggregation before they can be linked to the observational data.
% Here: Describe aims of the study or redundant with part 2? Something like this: The aim of our study is to investigate these data-related issues by synthesizing trait data from different databases.

2. We describe discrepancies in the definitions of traits used to create freshwater invertebrate trait databases in Europe, North America, New Zealand, and Australia. Based on our comparisons of these trait databases, we established 4 novel trait datasets by harmonising 7 commonly used grouping features. Two of these datasets were used to compare aggregated traits obtained by different aggregation methods with traits assigned by experts at family-level. We tested trait aggregation based on the mean and based on the median, each with different weightings. Furthermore, we explored the effect of grouping feature harmonisation and trait aggregation by re-analysing data from a case study.

3. Between the trait databases, grouping features differed in terms of their classification into traits and their scope leading to discrepancies how traits are defined. Also, the coding to describe traits varied between databases. Aggregated and assigned traits based on expert knowledge displayed a good match, especially approaches using the median. Grouping feature harmonisation and trait aggregation only slightly changed the results in the re-analysis compared to the original study, although averaging effects can occur due to harmonisation.

@@

Explaining and predicting how communities are shaped by environmental factors is one of the main goals of ecology. Species traits are measurable properties of an organism (\cite{mcgill_rebuilding_2006}), and comparing communities based on the traits possessed by species may facilitate testing of a range of ecological hypotheses (\cite{heino_jani_macroecological_2013}). Traits are assumed to be adaptations (e.g., physiological, behavioural, etc.) of organisms to their environment and represent direct or indirect linkages between the biological response of an organism or a population and its environment (\cite{southwood_habitat_1977, verberk_delivering_2013}).
Besides providing more mechanistic-based expressions of species-environment relationships, trait-based approaches might be suitable for large-scale analysis because the variability in trait responses is lower than in taxonomic responses (\cite{baird_toward_2011, bonada_taxonomic_2007}).

Traits of individual freshwater invertebrate are difficult to determine because - unlike plant traits - they are often difficult to measure directly. For example, to describe feeding habits requires that we understand both mouthpart morphology and the specific food that is consumed, and the organisms function within its community (\cite{moog_comprehensive_nodate}). Nevertheless, invertebrate traits have been increasingly used in freshwater ecology, e.g. by relating macroinvertebrate trait composition to environmental factors and as trait metrics in biomonitoring (\cite{bhowmik_large_2015, menezes_beyond_2010, poff_developing_2010, szocs_effects_2014}).

In the last decades, freshwater ecologists have compiled comprehensive invertebrate trait databases for various continents (\cite{kefford_integrated_2020, Philips_and_Smith_NZ_DB_2018, schmidt-kloiber_www.freshwaterecology.info_2015, tomanova_trophic_2006, ussegliopolatera_biological_2000, vieira_database_nodate}). The availability of invertebrate trait data from different continents arguably enables comparisons of trait variation and testing if variation in trait structure is consistently related to environmental factors across both small and large spatial extents. To date, such analyses have been carried out mostly within continents, using information from one or two trait databases. For example, \citet{bonada_taxonomic_2007} compared trait composition for Mediterranean and temperate regions in Europe based on traits from \citet{ussegliopolatera_biological_2000} (typically referred to as the Tachet database), \citet{poff_developing_2010} characterized trait composition across sites in the Western USA based on traits from \citet{poff_functional_2006}, and \citet{botwe_effects_2018} used trait definitions from \citet{poff_functional_2006} and \citet{schafer_trait_2011} to test for effects of salinity on invertebrate traits across different sites in South Australia. Analyses of invertebrate traits that synthesize information on invertebrate grouping features from more than two different continents are rare.

@@

Traits can be described in a binary fashion, or with multiple categories, ignoring uncertainty in how traits are expressed in any particular organism (e.g. adult terrestrial stage, presence of gills) or continuous (e.g. tolerance of pollution, body size). One approach for dealing with uncertainty is the use of fuzzy coding, where the traits are assigned probabilistic values. Fuzzy codes are used to account for plasticity in traits, variability in traits within taxonomic groups above species, and incomplete knowledge and are usually converted to proportions. In this study we denote the probabilistic values (0-1) of traits as affinities or affinity scores.

However, invertebrate trait databases are heterogeneous regarding the coding they use for their traits (\cite{culp_incorporating_2011}). For example, in the study from \citet{brown_functional_2018} the authors needed to reclassify traits using expert knowledge given that the individual databases employed different coding approaches (i.e. European and New Zealand databases fuzzy coding, North American database binary coding). Databases with standardised and unambiguous traits and a consistent coding of traits would minimize the data processing effort. Discrepancies in the taxonomic resolutions (e.g. species, genus or family) when linking observational taxonomic data to trait databases represents another challenge. When observations are at a more precise taxonomic level than data available in the trait databases (e.g. observations at species-level, trait data at genus-level) trait information of the less precise taxonomic level is often assigned (e.g. \cite{szocs_effects_2014, vos_taxonomic_2017}). Conversely, if trait information is only available at more precise taxonomic levels than the observed taxa, traits are aggregated to a less precise taxonomic level (e.g. \cite{aspin_extreme_2019, piliere_a._f._h._importance_2016, poff_functional_2006, szocs_effects_2014}). To data, studies have used different methods of trait aggregation, e.g. the mean (\cite{magliozzi_functional_2019}), median (\cite{szocs_effects_2014}) or the mode (\cite{piliere_a._f._h._importance_2016}). Studies on how and to which extent different trait aggregation methods influence trait-based analysis are missing. However, related knowledge would inform future studies regarding the choice of the aggregation method.

These challenges when working with or synthesizing trait data motivated us to (1) investigate discrepancies in trait definition between trait databases from Europe, North America, Australia, and New Zealand. As a basis for further evaluating different trait aggregation methods and how harmonising definition discrepancies between grouping features can influence trait-environment relationships, we developed 4 novel invertebrate trait datasets from these existing trait databases by harmonising 7 grouping features. (2) We then evaluated different trait aggregation methods by comparing aggregated traits with traits assigned by experts at family-level for two of the established trait datasets. (3) Finally, we re-analysed data from a study of salinisation effects on biological traits (\cite{szocs_effects_2014}) using data from the newly established harmonised European trait dataset and aggregated traits. By comparing the re-analysis's findings with those of \citet{szocs_effects_2014}, we investigated how harmonising and aggregating trait data can modify the outcome of trait-environment relationships.

@@

We extracted information from 6 trait databases from Europe, North America, Australia, and New Zealand and harmonised 7 grouping features. Trait information for Europe was obtained from the freshwaterecology.info database (\cite{schmidt-kloiber_www.freshwaterecology.info_2015}) and the Tachet database (\cite{ussegliopolatera_biological_2000}). The freshwaterecology.info contains taxa on species-level, while taxa recorded in the Tachet database are on species, genus and family-level. Trait information for North America was obtained from \citet{twardochleb_trait_data_2020} and complemented by \citet{vieira_database_nodate}. Data on body form for European and North American taxa was based on expert knowledge (\cite{polatera_personal_information_2020}). For Australia and New Zealand, we used trait databases from \citet{kefford_integrated_2020} and \citet{Philips_and_Smith_NZ_DB_2018} respectively. To increase readability we refer to the databases as well as the datasets we extracted from them by the name of the continent they originate from, except for European databases, which we refer to by their commonly used names (freshwaterecology.info and Tachet database).

We selected traits of seven grouping features that were available in all databases, are commonly applied in trait-based ecological studies, and describe different parts of the biology of a species: life history (Voltinism), morphology (Respiration, Body form, Size), ecology (Locomotion, Feeding mode) and reproduction (Oviposition). We omitted ecological traits that describe habitat preferences (e.g. temperature preference) because these traits are missing in the New Zealand trait database. Because grouping features were differently classified across the databases, we harmonised them into 26 traits (Table \ref{tab:traits_harmonisation}). Harmonisation was undertaken by amalgamating similar traits into one trait (e.g. crawlers and sprawlers into crawlers). When doing so, for each taxon the highest trait affinity score among the amalgamated traits was taken.

We used fuzzy coding for establishing our harmonised datasets unless data quality prohibited this. In the latter case, we used binary traits, i.e. categorical and continuous traits were converted into binary traits. For example, in the  freshwaterecology.info database, the classification of the trait voltinism accounts for different faunistic regions. Hence, entries such as "arctic" or "boreal" in the voltinism traits were substituted with a value of 1. Implicitly, we assumed for binary traits that a value of 1 and 0 corresponded to the highest and no affinity for a particular trait. Fuzzy codes are reported with different ranges in the trait databases (e.g. freshwaterecology.info 0 to 10, Tachet database 0 to 3 or 0 to 5). We standardised these to a range between 0 and 1 and converted trait affinities to percentages. Thus, fuzzy coded and binary traits were in the same range.

Prior to harmonisation, we consolidated duplicate taxa at the species, genus or family-level if present within a dataset, by either applying the median for fuzzy coded traits, or the maximum for binary traits. We also omitted all taxa with a lower taxonomic precision than family-level.

@@

    \textit{a} The trait "bluff (blocky)" occurred in the North America database (Vieira et al.) and was newly classified by expert knowledge into cylindrical and flattened (\cite{polatera_personal_information_2020}). \\  

@@

Aggregated trait affinities of the five trait aggregation methods (\textit{direct\_agg\textsubscript{median}}, \textit{direct\_agg\textsubscript{mean}}, \textit{stepwise\_agg\textsubscript{median}}, \textit{stepwise\_agg\textsubscript{mean}}, and \textit{weighted\_agg}) were compared to trait affinities assigned at family-level by experts, which were available for Australia and North America for a subset of grouping features and taxa. For the Australian dataset, we compared aggregated trait affinities with assigned trait affinities resolved at family-level for the grouping features feeding mode and size using data from \citet{chessman_dissolved-oxygen_2018}. In \citet{chessman_dissolved-oxygen_2018} feeding mode is classified similarly as in the Australian dataset except that the trait parasite is missing. We conducted the comparison for the 220 families available in \citet{chessman_dissolved-oxygen_2018}. Considering each factor combination of family and trait (in total 8) this amounted to 1760 cases.

@@

\subsection*{Effects of harmonisation and trait aggregation on trait-environment relationships}

We repeated the analysis of \citet{szocs_effects_2014} who studied the effect of anthropogenic salinization on invertebrates in the River Werra in Germany. For this re-analysis we used the established harmonised grouping feature dataset for Europe and additionally aggregated traits using the aforementioned aggregation methods.

The River Werra has been subject to effluents from the potash industry since the mid- 20th century and allows to study responses of invertebrates and their trait compositions to salinization (\cite{bathe_biological_2011}). Sites downstream, upstream, and close to the salt discharge (transition) were compared regarding their trait composition. Further details can be found in \citet{szocs_effects_2014}.

We substituted 6 of the grouping features from the original data with harmonised grouping features from the European trait dataset. We compared the results of the redundancy analysis (RDA) from the original study to the case when using harmonised grouping features. Specifically, the trait composition expressed as community weighted mean (CWM) traits was ordinated along an electric conductivity gradient. We compared the species scores obtained from the RDA, i.e. the coordinates of the tips of the vectors representing the CWM traits in the bi- or triplots. Following the original study, we identified traits associated with high or low salinity based on their distance to the ordination axis median using the Mahalanobis distance. CWM traits with a Mahalanobis distance greater than the $97.5 \%$ -quantile of the Chi-square distribution (5.02) were regarded as traits responding to either low or high salinity. For our analysis, we used the same 21 grouping features as in \citet{szocs_effects_2014}. The 6 harmonised grouping features used were \textit{Size, Feeding mode, Locomotion, Oviposition, Respiration}, and \textit{Voltinism}. Additionally, for testing the effect of aggregated traits we assigned to each taxon in \citet{szocs_effects_2014} the aggregated trait value for its corresponding family and repeated the RDA.

@@

The data processing and aforementioned analyses were carried out using R (Version 3.6.1). Raw data and the R code for data processing and grouping feature harmonisation are located in the Github repository: \url{https://github.com/KunzstLD/Invertebrate_traits}. Scripts and data to reproduce the trait aggregation and analysis with aggregated traits are located in the Github repository \url{https://github.com/KunzstLD/Trait-aggregation}.

@@

Regarding taxonomic coverage, the New Zealand dataset has the smallest taxon pool (478 taxa, Table \ref{tab:tax_coverage}). By contrast, the European trait dataset has the largest taxon pool with 4601 taxa followed by the North American dataset that contains trait information on 3753 taxa. The Australian dataset contains 1402 taxa. The European, New Zealand, and North American datasets contain most taxa at the species-level whereas the Australian dataset comprises a similar number of taxa at species and genus-level.

\begin{table}[ht]
    \centering
    \caption{Number of taxa per harmonised dataset and per taxonomic level. Numbers in parenthesis show rounded relative frequencies in percent.} 
    \label{tab:tax_coverage}
    \begin{tabular}{lccccc}
    \toprule[.1em]
    Dataset & Nr. of taxa & Species & Genus & Family & Nr. aquatic insects \\ 
    \toprule[.1em]
    EU & 4601 & 3739 (81) & 704 (15) & 158 (3) & 3942 (86) \\ 
    NOA & 3753 & 2414 (64) & 1163 (31) & 176 (5) & 3305 (88) \\ 
    AUS & 1402 & 564 (40) & 578 (41) & 260 (19) & 1016 (72) \\ 
    NZ & 478 & 404 (85) & 47 (10) & 27 (6) & 443 (93) \\ 
    \bottomrule
    \end{tabular}
\end{table}

@@

The amount of entries with available information for the selected grouping features varied strongly for the developed European, North American, and Australian datasets (Table \ref{tab:trait_coverage}). By contrast, the New Zealand dataset contained complete trait information for most of the investigated grouping features (between 94 \% and 100 \%).

\begin{table}[ht]
    \centering
    \caption{Rounded percentage of entries that include 
    information for the individual grouping features
    shown per trait dataset.} 
    \label{tab:trait_coverage}
    \begin{tabular}{lccccccc}
    \toprule[.1em]
    Dataset & Body form & Oviposition & Voltinism & Locomotion & Size & Respiration & Feeding mode \\ 
    \toprule[.1em]
    EU & 8 & 15 & 23 & 36 & 11 & 57 & 76 \\ 
    NOA & 28 & 13 & 47 & 52 & 73 & 44 & 63 \\ 
    AUS & 4 & 46 & 49 & 39 & 75 & 68 & 99 \\ 
    NZ & 100 & 94 & 100 & 99 & 100 & 100 & 99 \\ 
    \bottomrule
    \end{tabular}
\end{table}

@@

\subsection*{Discrepancies in invertebrate trait definitions across databases}

Definitions of grouping features and traits varied in their level of detail in the investigated trait databases. The freshwaterecology.info database, the Tachet database and the North American database (Twardochleb et al.) provided more detailed descriptions of their trait information compared to the North American (Vieira et al.) and New Zealand database. The Australian trait database (\cite{kefford_integrated_2020}) is a collection of seven specific trait datasets. Thus, grouping features occur multiple times with varying differentiation into traits. Depending on the dataset trait information is described with more or less detail.

The definition of grouping features varied across databases mainly concerning their differentiation into traits but also in their scope. We provide a summary of discrepancies in trait definitions in the appendix (Table \ref{stab:trait_definitions}). Both, differences in differentiation and scope can lead to discrepancies in trait definitions. For example, for the grouping feature feeding mode discrepancies arise because traits are assigned in different ways. The Tachet database defines predators as carvers, engulfers and swallowers. By contrast, in the North American (Twardochleb et al.) database predators are defined as engulfers and carnivorous piercers. In turn, in the Tachet database, piercers are defined as a separate trait encompassing herbivorous and carnivorous piercers. Furthermore, the scope in the freshwaterecology.info database for feeding mode is primarily on the food source of a species (except for filterers), while the other databases focus on the strategies of food acquisition. Therefore, the freshwaterecology.info database defines e.g. predator as "eating from prey", while the other databases use the mouthpart morphology as basis of their definition. The Tachet database captures the food source in an additional grouping feature. Varying levels of differentiation are also present in all other investigated grouping features between the trait databases (Table \ref{tab:trait_databases_coding_differentiation} and Table \ref{stab:trait_definitions}). Locomotion definitions differ also in scope between databases. Freshwaterecology.info and New Zealand databases describe locomotion as the way of movement of an organism, Tachet as substrate relation and locomotion, the North American (Vieira et al.) as how organisms deal with flow, Australia as attachment, and the North American (Twardochleb et al.) database includes not only the way of movement, but also the location of movement. Similarly, regarding reproduction traits, databases differ in their scope. Reproduction is captured in one grouping feature and defined as location of oviposit clutches and mode of reproduction in the freshwaterecology.info and Tachet databases. North America (Vieira et al.) provides information on the oviposition location but not on reproductive behavior. The Australian database report traits for reproductive behavior but also on oviposition site. The New Zealand database distinguishes three grouping features related to reproduction: reproductive technique, oviposition site (e.g. water surface, terrestrial), and egg/egg mass (e.g. free, cemented).

Various codings to describe traits are used throughout the databases (e.g. binary, fuzzy, continuous). The freshwaterecology.info and Australian databases use different codings, while the Tachet and the New Zealand database exclusively use fuzzy coding. Both North American trait databases contain categorical grouping features that can be converted into traits using a binary coding (Table \ref{tab:trait_databases_coding_differentiation}).

@@

We evaluated the simulation results based on the produced range of aggregated trait affinities and by comparing results of every simulated dataset.

As expected the produced range of trait affinities increased for all aggregation methods with increasing trait variability (Figure \ref{fig:overview_sim_results}).% Ben would add: 
 % as indicated by the increasing range of the boxes and wiskers with increasing standard deviation.
